---
title: "T-tests in R"
author: "Kareena del Rosario"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: lumen
    highlight: arrow
    toc: TRUE
---

### Prep the data

This code and corresponding dataset are available on github: https://github.com/kareena-delrosario/regression_rcode
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

p_needed <- c("tidyr",
          "tidyverse",
          "dplyr",
          "haven",
          "ggplot2",
          "readxl",
          "exact2x2",
          "purrr",
          "lsr")

packages <- rownames(installed.packages())
p_to_install <- p_needed[!(p_needed %in% packages)]

if(length(p_to_install) > 0){
  install.packages(p_to_install)
}

lapply(p_needed, library, character.only = TRUE)
```


```{r}
# read in data
data <- read_xlsx("/Users/kareenadelrosario/Desktop/local r code/regression_code/data_ttest.xlsx")

head(data)
```

### Create new variable that selects trust in democrats and republicans

```{r, warning=FALSE, message=FALSE}
# Option 1: similar to Python

## Trust in democrats
TrustD <- data[data$PARTY==1, "TRUST"]
## Trust in republicans
TrustR <- data[data$PARTY==2, "TRUST"]

# Option 2: dplyr

## Create "trustD" for Democrats
trustD <- data %>% 
  filter(PARTY == 1) %>%
  select(TRUST)

## Create "trustR" for Republicans
trustR <- data %>%
  filter(PARTY == 2) %>%
  select(TRUST)
  
```

------

## *Question*: Do Democrats trust in science (above chance, if chance is 75)?


### Preview the data


#### Base R: hist()


```{r, warning=FALSE, message=FALSE}
# simple histogram - plots frequency
hist(TrustD$TRUST)
```

#### ggplot


```{r, warning=FALSE, message=FALSE}
# similar to Python - plots density (scales to 1)
d_hist_plot <- trustD %>% 
  ggplot(aes(x = TRUST)) + 
  geom_histogram(aes(y = ..density..), fill = "cadetblue2", bins = 10, color = "black") +
  geom_density(color = "#327AAE") +
  geom_vline(xintercept = 75, color = "red",  size = 1) +
  labs(x = "Trust in Science", y = "Density") +
  ggtitle("Democrats Trust in Science") +
  theme_minimal()

print(d_hist_plot)
```

### Descriptives

We're going to shorten the mean and sd code to make our descriptives easier.

```{r}
mymean <- partial(mean, na.rm = TRUE)
mysd <- partial(sd, na.rm = TRUE)
```

```{r}
mymean(trustD$TRUST)
```

```{r}
mysd(trustD$TRUST)
```

```{r}
length(trustD$TRUST)
```

# One sample t-test

$$ t = \frac{\bar{x} - \mu_0}{s / \sqrt{N}}$$

```{r}
# manual calculation of t-test
# mu = population mean (75)

(88.59-75)/(9.59/sqrt(127))
```

Main function for t-tests

```{r one sample t tests}
# do Democrats trust in science
t.test(TrustD$TRUST, mu = 75) # can set specific population mean to test by setting mu
```

### Check whether TRUST is normally distributed


#### Base R: Q-Q Plot


```{r, warning=FALSE, message=FALSE}
qqnorm(trustD$TRUST)
```

#### ggplot


```{r}
# ggplot approach
ggplot(trustD, aes(sample = TRUST)) +
  stat_qq() +
  stat_qq_line(color = "red") + # add line to see normal dist
  ggtitle("Q-Q Plot of Democrat Trust") +
  theme_minimal()
```

#### Shapiro-Wilk Test

We can also quantitatively test whether the data are normally distributed.

- Democrats

```{r}
shapiro.test(trustD$TRUST)
```
- Republicans

```{r}
shapiro.test(trustR$TRUST)
```

#### Wilcoxon Test for non-normal dist

This is the non-parametric test for non-normally distributed data.

With continuity correction (default; treats like it's normally distributed)

```{r}
wilcox.test(trustD$TRUST)
```

Without continuity correction (only when you have super small samples)

```{r}
wilcox.test(trustD$TRUST, mu = 75, correct = FALSE)
```

----------------

# Two-samples t-test

### Visualize trust by political party

```{r}
bar_plot_party <- data %>%
  ggplot(aes(x = PARTY, y = TRUST, fill = PARTY)) + 
  stat_summary(fun=mean, geom="bar", position = position_dodge(), width = .3) + # plot mean as barplot
  stat_summary(fun.data = mean_se, geom = "errorbar", position = position_dodge(0.9), width = .1) + # mean_se is an argument that calculates SD (can calculate SE with additional arguments)
  labs(x = "Party", y = "Trust") + # add title to x and y-axis (labs = "labels")
  theme_minimal() + # preset theme
  ggtitle("Trust by Political Party")

bar_plot_party
```

##### Quick fix for categorical variables that are coded as continuous

```{r}
bar_plot_party <- data %>%
  ggplot(aes(x = factor(PARTY), y = TRUST, fill = factor(PARTY))) + 
  stat_summary(fun=mean, geom="bar", position = position_dodge(), width = .3) + # plot mean as barplot
  stat_summary(fun.data = mean_se, geom = "errorbar", position = position_dodge(0.9), width = .1) + # mean_se is an argument that calculates SD (can calculate SE with additional arguments)
  labs(x = "Party", y = "Trust in Science") + # add title to x and y-axis (labs = "labels")
  theme_classic() + # to match python
  ggtitle("Trust by Political Party")

bar_plot_party
```

##### Match python plot

```{r}
bar_plot_party +
    scale_x_discrete(labels=c('Democrats', 'Republicans', 'Independents')) + # x-axis labels
  theme(legend.position = "none") + # no legend
  ylim(0, 100) + # change y-axis "limits" to 0 to 100
  scale_fill_manual(values = c("#377eb8", "#e41a1c", "#ffcc33")) # change colors
```

#### Histogram

```{r, warning = FALSE}
# we're just going to plot democrats and republicans
data$PartyLabel <- ifelse(data$PARTY == 1, "Democrats", "Republicans") # assign a label

p <- ggplot(data, aes(x = TRUST, fill = PartyLabel)) +
  geom_histogram(aes(y = ..density..), position = "identity", alpha = 0.5, bins = 40) +
  geom_density(alpha = 0.3) + # transparency of density graphs
  scale_fill_manual(values = c("Democrats" = "#327AAE", "Republicans" = "#F93F17")) + # colors
  labs(x = "Trust in Science", y = "Density", fill = "Party") + # labels
  theme_minimal() +
  guides(fill = guide_legend(title = "Party")) + # title for legend
  ggtitle("Distribution of Trust in Science among Dems and Reps") +
  scale_x_continuous(limits=c(0,130)) + # change x-axis to match python
  scale_y_continuous(limits=c(0,.075)) # change y-axis (alternative to ylim)

print(p)

## If you want to save the image. It'll output to your working directory
# ggsave("trust.png", plot = p, dpi = 300, width = 8, height = 6)
```

#### Separate by party using facet_wrap()

Note that they share the same y-axis scale (unlike Python)

```{r, warning = FALSE}
p + facet_wrap(~PartyLabel)
```

--------------

# Independent samples t-test

- run an independent sample ttest (between subjects t-test) testing for differences in trust in science between Republicans and Democrats

```{r two sample t tests}
# Option 1 - default welch correction
t.test(trustD$TRUST, trustR$TRUST, var.equal = FALSE) # T Test comparing Group 1 to Group 2
```

```{r}
# get rid of party 3 (independents)
df_no_ind <- data %>% 
  filter(PARTY != 3)

# Option 2 - T-test without welch correction (same as Python)
t.test(TRUST ~ PARTY, data=df_no_ind, var.equal = T) 

```

```{r}
# package for effect size
library(lsr)

cohensD(TRUST ~ PARTY,
       data = df_no_ind) # reference data without independents
```

### Check assumptions of normality

#### Visualize with Q-Q Plot

```{r}
# ggplot approach
ggplot(df_no_ind, aes(sample = TRUST)) +
  stat_qq() +
  stat_qq_line(color = "red") + # add line to see normal dist
  ggtitle("Q-Q Plot of Trust") +
  theme_minimal() +
  facet_wrap(~ PartyLabel) # separate by party
```

#### Test with Shapiro

```{r}
# going back to our df with just democrats
shapiro.test(trustD$TRUST)
```
```{r}
# just republicans
shapiro.test(trustR$TRUST)
```
*How would you interpret the Shapiro test?*

### Mann–Whitney U test

FYI Main difference between this test and the Wilcoxon:

The Mann-Whitney U test (or Wilcoxon rank-sum test) is for two independent samples, while the Wilcoxon signed-rank test is for paired or matched samples.

```{r}
# not necessary to add the paired argument. Just want to make it clear that this is independent.
wilcox.test(trustD$TRUST, trustR$TRUST, paired = FALSE) 
```

-------------

# Paired samples t-test

We need the pre and post measurement in long format (i.e., same variable). We did this before, remember? How do we do that?

HINT: the columns we want to 'pivot' are Pre1 and Post1. We're creating a new column called TimePoint (should have Pre1 and Post1 names) and its values should be in a new column called Belief.

```{r}
long_df <- data %>% 
  pivot_longer(cols = c("Pre1", "Post1"),
               names_to = "TimePoint", 
               values_to = "Belief")

# check our work
long_df %>% 
  select(`Response ID`, TimePoint, Belief)
```

#### Visualize belief by timepoint

```{r, warning = FALSE}
bar_plot_paired <- long_df %>%
  ggplot(aes(x = TimePoint, y = Belief, fill = TimePoint)) + 
  stat_summary(fun=mean, geom="bar", position = position_dodge(), width = .3) + # plot mean as barplot
  stat_summary(fun.data = mean_se, geom = "errorbar", position = position_dodge(0.9), width = .1) + #
  labs(x = "Time Point", y = "Belief") + # add title to x and y-axis (labs = "labels")
  theme_classic() +
  ylim(c(0,80))

bar_plot_paired
```


```{r paired t test}
# using original data where pre and post are separate variables
t.test(data$Pre1, data$Post1, paired=T) # tests whether difference is significantly different from 0
```
Let's manually calculate Cohen's D, borrowing from Madalina's Python code. I'm simplifying it because we just need Cohen's D.

```{r}
cohen <- function(x1, x2) {
   xd = x1 - x2
   Md = mymean(xd)
   sd = mysd(xd)
   d = abs(Md) / sd
   
   print(d)
}

cohen(data$Pre1, data$Post1)
```

#### Wilcoxon Test for paired data

```{r}
wilcox.test(data$Pre1, data$Post1, paired = TRUE) 
```
#### Descriptives for reporting

You can get the means and sds the old-fashioned way, but if you plan to get descriptives from several variables, you might as well build a function.

```{r}
descriptives <- function(x1, x2) {
  m1 <- mymean(x1)
  m2 <- mymean(x2)
  s1 <- mysd(x1)
  s2 <- mysd(x2)
  
 # Create a data frame for nicer display - could use table() but it's ugly
  results_df <- data.frame(
    Metric = c("Mean", "SD"), # 3 columns = metric, pre, post
    Pre1 = c(m1, s1),
    Post1 = c(m2, s2)
  )
  
  return(results_df)
}

descriptives(data$Pre1, data$Post1)
```
